---
title: "Rendering to Folder"
author: "Bruce Cochrane"
date: "February 19, 2015"
output: html_document
---
We want to develop a function to render markdown-created html to a particular folder, and then perhaps develop an lapply function to do all of the conversions.
```{r}
library(rmarkdown)
```
Now see if we can render something without designating the package.

render ("./Rmd/file",output_dir="../html")

An example
```{r}
render("./Rmd/BasicR.Rmd",output_dir="../html")
```
So now a function to mass render everything (no doubt will have errors)

### Functionalize the above
```{r}
rendHtml <-function(file){
  f.name <-paste("./Rmd/",file,".Rmd",sep="")
  rmarkdown::render(f.name,output_dir="../html")
}
```
OK.  Now to extract a list of all of the .Rmd files in the folder
```{r}
fls <-list.files("./Rmd")
fls
```
Now, we want to extract those that end with .Rmd  Looks like we can do this pretty easily.

```{r}
fls.rmd <-fls[grep("*.Rmd",fls)]
fls.rmd
```
How cool.  Now let's test the function on an easy one first.

```{r}
rendHtml("TOC")
```
So now, can we use lapply safely?  We'll do it for a subset first
```{r}
length(fls.rmd)
```
OK.  44 files.  Let's try the first five.  But before we can do that, we have to remove the suffixes.  From nabble, the basic form is

substr((myfile), 1, nchar((myfile)) - 4) 
```{r}
fls.rmd.nm <-sapply(fls.rmd, function (x) substr(x,1,nchar(x)-4))
names(fls.rmd.nm)<-NULL
fls.rmd.nm
```
OK.  Now do the first five
```{r}
sapply(fls.rmd.nm[22:25],rendHtml)
```
One nasty thing we learned - with .Rmd files in their own directory, path begins there (not with the working directory).  I wonder if we could fix that by setting the path initially in the file.

Enough said.  We need to have absolute addresses to the data.  That's going to take more work than I want to do today.
