---
title: <h1>Comparing Estimators - The Tajima Test</h1>
date:  <center>`r format(Sys.Date(),format="%B %d %Y")`</center>
bibliography:  ../TPG.bib
csl: ../genetics.csl.txt
output:
  html_document:
    css: ../Chapter.css
    toc: yes
    toc_depth: 5
    code_fold: show
---


## The Tajima Statistic (D)

Now that we have our two estimators, we need to compare them and then somehow determine if they differ by more than we would expect by chance.  This brings us to one of the most widely  used statistics in molecular population genetics - Tajima's D.  What we will see is that, while the calculation of the estimator is straightforward, its distribution is another one (like the Ewens estimator) that is best determined by simulation. In addition, it is important to note at the outset is that there is not necessarily a single explanation for an outcome that differs significantly from expectation.   What we will do first, therefore, is to define the measure and then look at some real data.

### The Formula

The easiest way to see what is going on is to look at the formula (9.2 in @nielsen2013introduction):

\(\large{D=\frac{\hat{\theta}_T - \hat{\theta}_W}{\sqrt{\hat{V}(\hat{\theta}_T - \hat{\theta}_W)}}}\)



In this formula, \(\theta_T\) is the notation used for π and  \(\theta_W\) is the Watterson estimator, based as we've seen previously on the number of segregating sites. the denominator is the square root of the variance of the difference between the two estimators; it is thus always positive.  The numerator, on the other hand can be either positive or negative.  It is positive when π is large, which we've already seen occurs when there are many intermediate frequency sites; it is small (and D negative) when there are lots of sites with only a few variants each.

### Some data

But enough of hypotheticals.  What we want to do is to look at actual data and calculate D.  Once we have, we'll then focus on interpreting the result.  For this, we will use popset [345105052](http://www.ncbi.nlm.nih.gov/popset/?term=345105052), from the *Drosophila mauritania* *spaghetti squash* gene, which were imported into Clustal.  A deletion-free segment was then saved to use in this example

```{r}
library(TeachingPopGen)
data(spaghetti)
dat <-spaghetti
dat
```

As before, we can plot the site frequency spectrum

```{r,warning=FALSE}
plot(site.spectrum(dat))
length(seg.sites(dat))
```

And we see that most of the `r length(seg.sites(dat))` segregating sites fall in the first class, with one sequence differing from all of the rest.  The function for the Tajima test is simply

```{r}
D <-tajima.test(dat)
D
```
We note that three values are returned:

1.  $D is the actual value of the statistic.  In this case we see that it is negative, suggesting that the tendency is towards a disproportionate number of low frequency variants.
2.  $Pval.normal and $Pval.beta are estimates of the probability of the data, based on the assumption of the distribution of D being either a normal or a beta distribution.  These values are approximations at best; we will see that simulation provides a much better means of comparing our results to an expected distribution.

### What does this tell us?

The negative value of D tells us that there are more low frequency sites than we would expect.  What does that suggest about the evolutionary history of the gene and species?  We will explore this further, but for now, we can think of a few possibilities:


1.  Most mutations are detrimental, so that they tend to be eliminated by selection, so that their frequencies are low (a negative selection model).
2.  One mutation has arisen recently and is advantageous, leading to its increase in frequency, and thus decrease in relative frequency of others
3.  The population has undergone a recent expansion, so that there are many more recent rather than ancient branches.  In that case, there would be many more mutations that would be represented only a few times in the sample, leading to an excess of low frequency sites.

The first two hypotheses invoke natural selection as an explanation, something we will explore in depth in the second part of the course.  The third does not - it suggests that in fact, patterns of sequence diversity may be signatures of past demographic events.  If so, then perhaps the opportunity exists to use data like these to make inferences about those historical processes.

### Conclusions

As noted, Tajima's D test is one of the most widely used statistics in molecular population genetics, however we've left two problems outstanding.  First, while the expected value under our mutation-drift model is zero (or close to it), we don't know a lot about its distribution.  Second, even if we see a departure from expectation, the alternative biological hypotheses are complex and diverse.  To explore these questions, in the next unit we will see how simulation can provide real insight.

