---
title: <center>rQTL</center>
author: <center>Bruce Cochrane</center>
date: <center>`r format(Sys.Date(),format="%B %d %Y")`</center>
output: 
    html_document:
       css: Chapter.css
       toc: yes
       toc_depth: 5
---


A quick and very rough introduction to qtl mapping in R.  

```{r}
library(qtl)
```
First, look at raw data

```{r}
dat.raw <-read.csv("http://www.rqtl.org/sug.csv")
dat.raw[1:8,1:10]
```

So we see that we have four phenotypes, sex, and then genotypes for a bunch of loci.  First two lines (not shown) have chromosome and position for each marker.

Now use the package to input this stuff in a more usable form

```{r}
sug <- read.cross("csv", "http://www.rqtl.org", "sug.csv",
                  genotypes=c("CC", "CB", "BB"), alleles=c("C", "B"))
summary(sug)
```
Note that sex and ID are treated as phenoytpes.

Now look at the genetic map:

```{r}
plot.map(sug)
```

Look at phenotype 1 (blood pressure:

```{r}
plotPheno(sug, pheno.col=1)
```

Do a single QTL analysis on phenotype 1 ()

```{r}
sug <- calc.genoprob(sug, step=1)

out.bp <- scanone(sug)

summary(out.bp)
plot(out.bp)
```

Go back and look for high probability genes

```{r}
summary(out.bp, threshold=3)
```


Do the same thing for heart rate

```{r}
out.hr <-scanone(sug,pheno.col = 2)
plot(out.hr)
summary(out.hr,threshold=3)
```

Now do phenotype 3 (birth weight)
```{r}
out.bw <-scanone(sug,pheno.col = 3)
plot(out.bw)
summary(out.bw,threshold=3)
```

So all three have a gene on chromosome 15, and in the latter two cases they are pretty close.  But the same gene?  Difficult to conclude, but absolutely need more detailed map.