---
title: "BIO 444 Test Questions"
output: html_notebook
---

# 1 Unix Loops
In order to run our migrate-n analyses we had to construct directories, each containing a parameter settings file, and then enter each directory to run migrate. Imagine that instead of 1 species, you had to analyze 40 species with migrate-n.

Write a 'for-loop' in unix that would allow you to start in a directory containing directories with all five hypotheses, move into each directory, start migrate-n, and then move to the next directory until migrate models are running in each directory.

Ignore the need to create virtual terminal windows (screen) for this problem: imagine that migrate-n starts in the background and then lets you issue more commands. To over come this when testing your program, use "echo migrate-n parmfile" as the command to run migrate-n.

You may use the directory linked [here](http://ericcrandall.github.io/lessons/exam/5_hypothesis.zip) to experiment with. Give the code in the box below.

```{bash}
for dir in *.*
  do
  echo dir
  cd dir
  echo migrate-n parmfile
  cd ..
done

```

# 2 Bat R Throw L
The following three questions will be based on this comma-delimited dataset that describes statistics for Major League Baseball in 2008. Please paste in your one line of code, and the resulting R output into the response box.

How many National League baseball players batted right and threw with their left in 2008?

```{r}
bat<-read.table("https://ericcrandall.github.io/BIO444/lessons/exam/batting_players08.txt", header=T, sep=",")

table(bat$bats,bat$throws,bat$lgID)
```
# 3 Names
What are their names? How many runs-batted-in (RBI) did they hit?

```{r}
BRTL<-bat[which(bat$bats=="R" & bat$throws=="L" & bat$lgID=="NL"),c("nameFirst","nameLast","RBI")]
```

# 4 RBI Table
Create a table that is sorted alphabetically by the number of RBIs these players hit in 2008.
```{r}
BRTL[order(BRTL[,3],decreasing=T),]
```

