---
title: Making Maps With R
author: "Eric D. Crandall"
output:
  html_document:
    toc: yes
  bookdown::html_chapter:
    toc: no
---



# Making Maps with R {#map-making-in-R} 

```{r, include = FALSE}
library(knitr)
opts_chunk$set(fig.width=10,
               fig.height=7,
               out.width = "600px",
               out.height = "420px",
               fig.path = "lecture_figs/making-maps-")
```
Thanks to Eric Anderson for ideas and some code and data in this lecture.

## Intro {#map-making-intro}

For a long time, R has had a relatively simple mechanism, via the `maps` package, for making simple outlines
of maps and plotting lat-long points and paths on them.

With the advent of packages like `sp`, `rgdal`, and `rgeos`, R has been acquiring much of the
functionality of traditional GIS packages (like ArcGIS, etc). These packages require
installation of external libraries and considerable familiarity with GIS concepts, and are
not as accessible as they might be (I don't use them).

More recently, a third approach to convenient mapping, using `ggmap` has been developed that
allows the tiling of 
detailed base maps from Google Earth or Open Street Maps, upon which spatial data may be
plotted.
We are going to focus on mapping using base maps from R's tried and true `maps`
package and also using the
`ggmap` package.

As with the graphics lesson, we are going to completely skip over R's base graphics system and head directly to Hadley Wickham's `ggplot2` package.  Hadley has
included a few functions that make it relatively easy to interact with the data in R's
`maps` package, and of course, once a map layer is laid down, you have all the power of
ggplot at your fingertips to overlay whatever you may want to over the map.  `ggmap` is a
package that goes out to different map servers and
grabs base maps to plot things on, then it sets up the coordinate system and writes it out
as the base layer for further ggplotting.  It's nice, but does not support different projections.


For today we will also be skipping how to read in traditional GIS "shapefiles" so as to
minimizethe number of packages that need installation, but keep in mind that it isn't too hard to do that in R, too.

### Prerequisites
You are going to need to install a few extra packages to follow along with this lecture.
```{r, eval=FALSE}


# some standard map packages.
install.packages(c("maps", "mapdata", "ggpmap"))

# the github version of ggmap, JIK
#devtools::install_github("dkahle/ggmap")
```




### Load up a few of the libraries we will use

```{r}
library(ggplot2)
library(ggmap)
library(maps)
library(mapdata)
library(dplyr)
```


## Plotting maps-package maps with ggplot  {#maps-package-and-ggplot}

### The main players:

* The `maps` package contains a lot of outlines of continents, countries, states, and counties that have
been with R for a long time.  
* The `mapdata` package contains a few more, higher-resolution outlines.
* The `maps` package comes with a plotting function, but, we will opt to use `ggplot2` to plot the 
maps in the `maps` package.  
* Recall that `ggplot2` operates on data frames.  Therefore we need some way to
translate the `maps` data into a data frame format the `ggplot` can use.

### Maps in the maps package

* Package `maps` provides lots of different map outlines and points for cities, etc.  
* Some examples: `usa`, `nz`, `state`, `world`, etc.



### Makin' data frames from map outlines

* `ggplot2` provides the `map_data()` function.
    + Think of it as a function that turns a series of points along an outline into a data frame
    of those points.
    + Syntax:  `map_data("name")` where "name" is a quoted string of the name of a map in the `maps` or `mapdata`
    package
* Here we get a USA map from `maps`:
    ```{r}
    usa <- map_data("usa")

    dim(usa)
    
    head(usa)
    
    tail(usa)
    ```
* Here is the high-res world map centered on the Pacific Ocean from `mapdata`
    ```{r}
    w2hr <- map_data("world2Hires")

    dim(w2hr)

    head(w2hr)

    tail(w2hr)
    ```

### The structure of those data frames
These are pretty straightforward:

* `long` is decimal-degree longitude.  Things to the west of the prime meridian are negative.
* `lat` is decimal-degree latitude.
* `order`. This just shows in which order `ggplot` should "connect the dots"
* `region` and `subregion` tell what region or subregion a set of points surrounds.
* `group`.  This is _very important_!  `ggplot2`'s functions can take a group argument which 
controls (amongst other things) whether adjacent points should be connected by lines.  If they are
in the same group, then they get connected, but if they are in different groups then they don't.
    + Essentially, having to points in different groups means that `ggplot` "lifts the pen" when going between them.
    

### Plot the USA map

* Maps in this format can be plotted with the polygon geom.  i.e. using `geom_polygon()`.
* `geom_polygon()` drawn lines between points and "closes them up" (i.e. draws a line from the last
point back to the first point)
* You have to map the `group` aesthetic to the `group` column
* Of course, `x = long` and `y = lat` are the other aesthetics.

#### Simple black map
By default, `geom_polygon()` draws with no line color, but with a black fill:
```{r}
usa <- map_data("usa") # we already did this, but we can do it again
ggplot() + geom_polygon(data = usa, aes(x=long, y = lat, group = group)) + 
  coord_fixed(1.3)
```
Boink!
#### What is this coord_fixed()?

* This is very important when drawing maps.
* It is part of the `coord` series of functions in ggplot2 that we didn't cover last time.
* It fixes the relationship between one unit in the $y$ direction and one unit in the $x$ direction.
* Then, even if you change the outer dimensions of the plot (i.e. by changing the window size or the size
of the pdf file you are saving it to (in `ggsave` for example)), the _aspect ratio_ remains unchanged.
* In the above case, I decided that if every $y$ unit was 1.3 times longer than an $x$ unit, then  the
plot came out looking good.
    + A different value might be needed closer to the poles.
```{r}
ggplot() + geom_polygon(data = usa, aes(x=long, y = lat, group = group)) + 
  coord_fixed(4)
ggplot() + geom_polygon(data = usa, aes(x=long, y = lat, group = group)) + 
  coord_fixed(0.5)
```
    
#### Mess with line and fill colors

* Here is no fill, with a red line.  Remember, fixed value of aesthetics go _outside_ the `aes` function.
    ```{r}
    ggplot() + 
      geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = NA, color = "red") + coord_fixed(1.3)
    ```
    
* Here is violet fill, with a blue line.
    ```{r}
    gg1 <- ggplot() + 
      geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = "violet", color = "blue") +  coord_fixed(1.3)
    gg1
    ```
    
#### Adding points to the map

* Let's add black and yellow points at CSUMB, the White House, and Toad Suck, Arkansas.
    ```{r}
    points <- data.frame(
      long = c(-121.8018, -77.0365, -92.5599),
      lat = c(36.6544, 38.8977, 35.0756),
      names = c("CSUMB", "White House", "Toad Suck"),
      stringsAsFactors = FALSE
      )  

    gg1 + 
      geom_point(data = points, aes(x = long, y = lat), color = "black", size = 5) +
      geom_point(data = points, aes(x = long, y = lat), color = "yellow", size = 4)
    ```

#### See how important the group aesthetic is

Here we plot that map without using the group aesthetic:
```{r}
ggplot() + 
      geom_polygon(data = usa, aes(x=long, y = lat), fill = "violet", color = "blue") + 
      geom_point(data = points, aes(x = long, y = lat), color = "black", size = 5) +
      geom_point(data = points, aes(x = long, y = lat), color = "yellow", size = 4) +
      coord_fixed(1.3)
```

That is no bueno!  The lines are connecting points that should not be connected! That's because its assuming that everything is one group, and not "lifting the pen."



### State maps
We can also get a data frame of polygons that tell us above state boundaries:
```{r}
states <- map_data("state")
dim(states)

head(states)

tail(states)
```

#### Plot all the states, all colored a little differently

This is just like it is above, but we can map fill to `region` and 
make sure the the lines of state borders are white.
```{r}
ggplot(data = states) + 
  geom_polygon(aes(x = long, y = lat, fill = region, group = group), color = "white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE)  # do this to leave off the color legend
```

Boom! That is easy.

#### Plot just a subset of states in the contiguous 48:

* Read about the `subset` command.  It provides another way of 
subsetting data frames (sort of like using the `[ ]` operator with
a logical vector).
* We can use it to grab just CA, OR, and WA:
    ```{r}
    best_coast <- subset(states, region %in% c("california", "oregon", "washington"))
    
    ggplot(data = best_coast) + 
      geom_polygon(aes(x = long, y = lat), fill = "palegreen", color = "black") 
    ```

#### Man that is ugly!! 

* What have we forgotten here?
    
    ```{r}
    ggplot(data = best_coast) + 
      geom_polygon(aes(x = long, y = lat, group = group), fill = "palegreen", color = "black") + 
      coord_fixed(1.3)
    ```

Phew! That is a little better!


#### Zoom in on California and look at counties

* Getting the California data is easy:
    ```{r}
    ca_df <- states[states$region == "california",]

    head(ca_df)
    ```

* Now, let's also get the county lines there
    ```{r}
    counties <- map_data("county")
    ca_county <- counties[counties$region == "california",] 

    head(ca_county)
    ```

* Plot the state first but let's ditch the axes gridlines, and gray background by
using `theme_nothing()`.
    ```{r}
    ca_base <- ggplot(data = ca_df, mapping = aes(x = long, y = lat, group = group)) + 
      coord_fixed(1.3) + 
      geom_polygon(color = "black", fill = "gray")
    ca_base + theme_nothing()
    ```

* Now plot the county boundaries in white:
    ```{r}
    ca_base + theme_nothing() + 
      geom_polygon(data = ca_county, fill = NA, color = "white") +
      geom_polygon(color = "black", fill = NA)  # get the state border back on top
    ```

#### Get some facts about the counties


```{r, warning=FALSE, message=FALSE}
pop_and_area<-read.csv(file="./data/ca_counties.csv")
      
    ```
* We need to attach those to every point on polygons of the counties.  This is a job for `inner_join` from the `dplyr` package
    ```{r}
    cacopa <- inner_join(ca_county, pop_and_area, by = "subregion")
    ```
* And finally, add a column of `people_per_mile`:
    ```{r}
    cacopa$people_per_mile <- cacopa$population / cacopa$area

    head(cacopa)
    ```

#### Now plot population density by county

If you were needing a little more elbow room in the great Golden State, this shows you where you can find it. This also gives us a chance to learn about theme(), which is the par() of ggplot2, and I forgot to show you earlier. 
  * It allows control over all the non-data aesthetics of your plot. 
  * `element_blank()` is what we use to turn off items.
  * You can read more in the `?theme` man page about it.

Since we still want the guides and legends, we can't use `theme_nothing` here.
```{r}

ditch_the_axes <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank()
  )

elbow_room1 <- ca_base + 
      geom_polygon(data = cacopa, aes(fill = people_per_mile), color = "white") +
      geom_polygon(color = "black", fill = NA) +
      theme_bw() +
      ditch_the_axes

elbow_room1 
```

#### Lame!

* The popuation density in San Francisco is so great that it makes it hard to discern differences between
other areas.
* This is a job for a scale transformation.  Let's take the log-base-10 of the population density.
* Instead of making a new column which is log10 of the `people_per_mile` we can just apply the
transformation in the gradient using the `trans` argument
```{r}
elbow_room1 + scale_fill_gradient(trans = "log10")
```

 
#### Still not great
I personally like more color than ggplot uses in its default gradient.
```{r}
eb2 <- elbow_room1 + 
    scale_fill_gradientn(colours = rev(rainbow(7)),
                         breaks = c(2, 4, 10, 100, 1000, 10000),
                         trans = "log10")
eb2
```

That is reasonably cool.



### zoom in?
Note that the scale of these maps from package `maps` are not great. We can zoom in to the 
Bay region, and it sort of works scale-wise, but if we wanted to zoom in more, it would
be tough.  

Let's try!
```{r}
eb2 + xlim(-123, -121.0) + ylim(36, 38)
```

* Whoa! That is an epic fail. Why?
* Recall that `geom_polygon()` connects the end point of a `group` to its starting point.
* And also remember: the `xlim` and `ylim` functions in `ggplot2` discard all the data that is
not within the plot area.  
    + Hence there are new starting points and ending points for some groups (or in this case the
    black-line permiter of California) and those points get connected.  Not good.


### True zoom.

* If you want to keep all the data the same but just zoom in, you can use the `xlim` and `ylim` arguments to `coord_fixed()`. 
* This chops stuff off but doesn't discard it from the data set:
    ```{r}
    eb2 + coord_fixed(xlim = c(-123, -121.0),  ylim = c(36, 38), ratio = 1.3)
    ```


## ggmap {#ggmap-hooray}

The `ggmap` package is lots of fun.  You might be able to get
better looking maps at some resolutions by using shapefiles and rasters from naturalearthdata.com
but `ggmap` will get you 95% of the way there with only 5% of the work!

### Three examples

* I am going to run through three examples.  Working from the small spatial scale up to a larger spatial scale.
    1. Named "sampling" points on the Sisquoc River from the "Sisquoctober Adventure"
    2. A GPS track from a short bike ride in Wilder Ranch.
    3. Fish sampling locations from the coded wire tag data base.
    
### How ggmap works

* ggmap simplifies the process of downloading base maps from Google or Open Street Maps or Stamen Maps
to use in the background of your plots.
* It also sets the axis scales, etc, in a nice way.  
* Once you have gotten your maps, you make a call with `ggmap()` much as you would with `ggplot()`
* Let's do by example.

### Sisquoctober

* Here is a small data frame of points from the Sisquoc River.
    ```{r}
    sisquoc <- read.table("data/sisquoc-points.txt", sep = "\t", header = TRUE)
    sisquoc

    ```
  * ggmap typically needs a zoom level:
    + Zoom levels go from 3 (world scale) to 20 (house scale)
    ```{r}
    # compute the mean lat and lon so we can center the map on it
    ll_means <- sapply(sisquoc[2:3], mean)
    sq_map2 <- get_map(location = ll_means,  maptype = "satellite", source = "google", zoom = 15)
    ggmap(sq_map2) + 
      geom_point(data = sisquoc, color = "red", size = 4) +
      geom_text(data = sisquoc, aes(label = paste("  ", as.character(name), sep="")), angle = 60, hjust = 0, color = "yellow")
    ```
* That's decent.  How about if we use the "terrain" type of map:
    ```{r}
    sq_map3 <- get_map(location = ll_means,  maptype = "terrain", source = "google", zoom = 15)
    ggmap(sq_map3) + 
      geom_point(data = sisquoc, color = "red", size = 4) +
      geom_text(data = sisquoc, aes(label = paste("  ", as.character(name), sep="")), angle = 60, hjust = 0, color = "yellow")
    ```

### How about a bike ride?

* We can plot a route like this:
    ```{r}
    bike <- read.csv("data/bike-ride.csv")
    head(bike)
    bikemap1 <- get_map(location = c(-122.080954, 36.971709), maptype = "terrain", source = "google", zoom = 14)
    ggmap(bikemap1) + 
      geom_path(data = bike, aes(color = elevation), size = 3, lineend = "round") + 
      scale_color_gradientn(colours = rainbow(7), breaks = seq(25, 200, by = 25))
    ```
* See how we have mapped elevation to the color of the path using our
rainbow colors again.
* Note that getting the right zoom and position for the map is sort of trial and
error.  You can go to google maps to figure out where the center should be (right click and choose "What's here?" to get the lat-long of any point. )

### Pelagic Fishes
Let's finally get a visual for the data in the pelagics metadata set that we've been playing with, shall we?
```{r}
pelagics<-read.table("./data/pelagics_metadata.txt",sep="\t",header=T,stringsAsFactors = F)
#Instead of playing with the scale to get it just right, I'm going to try make_bbox to make
#a bounding box
boxtest <- make_bbox(lat=decimalLatitude,lon=decimalLongitude,data=pelagics)
pmap <- get_map(location = boxtest, maptype = "satellite", source = "google")

ggmap(pmap) + geom_point(data = pelagics, mapping = aes(x = decimalLongitude, y = decimalLatitude, color = paste(pelagics$genus,pelagics$species))) +
  theme(legend.position="bottom") + 
  labs(color="Species")

# Huh, it chopped off a whole bunch of data for some reason. Can't get it to fix either. Going back to  zoom
llmean<-sapply(pelagics[,c("decimalLongitude","decimalLatitude")],mean)

pmap <- get_map(location = llmean, maptype = "satellite", source = "google",zoom=4)

ggmap(pmap) + geom_point(data = pelagics, mapping = aes(x = decimalLongitude, y = decimalLatitude, color = paste(pelagics$genus,pelagics$species))) +
  theme(legend.position="bottom") + 
  labs(color="Species")
```
It is still chopping off Aceh! Also it would be nice to make pie charts out of the species to show their relative frequency at each location. But that is for another R session. It is almost midnight!

#messing with vote data

```{r}
votes<-read.table(file="./data/state_votes.txt",sep="\t",header=T,row.names=T)
votes$State.1<-tolower(votes$State.1)
{r}
votes2<-votes[,44:58]
votes3<-t(votes2)
write.csv(votes3,"./data/state_votes.txt")


votes<-read.csv(file="./data/state_votes.txt", stringsAsFactors=F)

length(votes$Alabama[which(votes$Alabama=="R")])/length(votes$Alabama)

voteprop<-function(x){
  length(x[which(x=="R")])/length(x)
}

Rs<-sapply(votes,voteprop)


Rs2<-as.data.frame(Rs)
Rs2$region<-tolower(rownames(Rs2))
states <- map_data("state")

states2<-inner_join(states,Rs2,by="region")

ggplot(data = states2) + 
  geom_polygon(aes(x = long, y = lat, fill = Rs, group = group), color = "white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE) + scale_fill_continuous(low = "Blue", high = "Red")

```
